Joint estimation of ancestral states and phylogenetic model parameters is inconsistent

### 2017-11-21 appendix.tex
@.@

\begin{table}
\centering
\begin{tabular}{|l|l|l|}
    \hline
$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
    \hline
    $\emptyset$    &$1+x'^2+y'^2+4x'y'w'+x'^2y'^2$&$1+2x'y'+2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2$\\
    $\{1\}$        &$1-x'^2+y'^2-x'^2y'^2$&$1-x'^2w'+y'^2w'-x'^2y'^2$\\
    $\{2\}$        &$1+x'^2-y'^2-x'^2y'^2$&$1+x'^2w'-y'^2w'-x'^2y'^2$\\
    $\{3\}$        &$1-x'^2+y'^2-x'^2y'^2$&$1-x'^2w'+y'^2w'-x'^2y'^2$\\
    $\{1,2\}$      &$1-x'^2-y'^2+x'^2y'^2$&$1+2x'y'-2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2$\\
    $\{1,3\}$      &$1+x'^2+y'^2-4x'y'w'+x'^2y'^2$&$1-2x'y'-2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2$\\
    $\{2,3\}$      &$1-x'^2-y'^2+x'^2y'^2$&$1-2x'y'+2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2$\\
    $\{1,2,3\}$    &$1+x'^2-y'^2-x'^2y'^2$&$1+x'^2w'-y'^2w'-x'^2y'^2$\\
    \hline
\end{tabular}
\caption{Site pattern probabilities.
All values are multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
    \multicolumn{3}{c}{$\tau=\tau_1$}\\
    \hline
    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
    \hline
    $\emptyset$&
    $\emptyset$&
    $(1+x')^2   (1+w')(1+y')^2$\\
     $\{1\}$    &
    $\emptyset$&
    $(1+x')(1-x')(1+w')(1+y')^2$\\
     $\{2\}$    &
    $\emptyset$&
    $(1+x')^2   (1+w')(1+y')(1-y')$\\
     $\{3\}$    &
    $\emptyset$&
    $(1+x')(1-x')(1+w')(1+y')^2$\\
    $\{1,2\}$  &
    $\{\emptyset,\{1,2\}\}$&
    $(1+x')(1-x')(1+w')(1+y')(1-y')$\\
    $\{1,3\}$  &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{1\}\\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1-x')^2   (1+w')(1+y')^2\\
                    (1+x')^2   (1-w')(1+y')^2\\
                    (1+x')^2   (1+w')(1-y')^2
                \end{array}$\\
    $\{2,3\}$  &
                $\{\emptyset,\{1,2\}\}$&
                $(1+x')(1-x')(1+w')(1+y')(1-y')$\\
    $\{1,2,3\}$&
                $\{1,2\}$&
                $(1+x')^2   (1+w')(1+y')(1-y')$\\
    \hline
    \multicolumn{3}{c}{$\tau=\tau_2$}\\
    \hline
    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
    \hline
    $\emptyset$       &$\emptyset$&$(1+x')^2   (1+w')(1+y')^2$\\
    $\{1\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                        (1+x')(1-x')(1+w')(1+y')^2\\
                        (1+x')^2   (1-w')(1+y')(1-y')
                    \end{array}$\\
      $\{2\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x')^2   (1+w')(1+y')(1-y')\\
                    (1+x')(1-x')(1-w')(1+y')^2
                    \end{array}$\\
      $\{3\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x')(1-x')(1+w')(1+y')^2\\
                    (1+x')^2   (1-w')(1+y')(1-y')
                    \end{array}$\\
     $\{1,2\}$         &
    $\left\{\begin{array}{l}
                    \{\emptyset,\{1,2\}\}\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x')(1-x')(1+w')(1+y')(1-y')\\
                    (1+x')^2   (1-w')(1+y')^2
                    \end{array}$\\
     $\{1,3\}$         &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{\{1\},\{2\}\}\\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1-x')^2   (1+w')(1+y')^2\\
                    (1+x')(1-x')(1-w')(1+y')(1-y')\\
                    (1+x')^2   (1+w')(1-y')^2
                    \end{array}$\\
      $\{2,3\}$        &
    $\left\{\begin{array}{l}
                    \{\emptyset,\{1,2\}\}\\
                    \{1\}\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x')(1-x')(1+w')(1+y')(1-y')\\
                    (1-x')^2   (1-w')(1+y')^2\\
                    (1+x')^2   (1-w')(1-y')^2
                    \end{array}$\\
     $\{1,2,3\}$       &
    $\left\{\begin{array}{l}
                    \{2\} \\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x')(1-x')(1-w')(1+y')^2 \\
                    (1+x')^2   (1+w')(1+y')(1-y')
                    \end{array}$\\
    \hline
\end{tabular}
\caption{Likelihood calculations for all site patterns and ancestral state partitions.
All values multiplied by $1/32$.
Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
See Appendix for full calculations.}
\label{tab:likelihoods}
\end{table}

@@

Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$.
Following \eqref{eq:hadamard_probability},

@@

p_{\emptyset} = P(\siteSplitRV=\emptyset \mid \tau=\tau_1,t=\{x,y,y\}),

@@

    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \nonumber \\
    &\qquad + p_{1}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \nonumber \\
    &\qquad + p_{2}          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \nonumber \\
    &\qquad + p_{3}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \nonumber \\
    &\qquad + p_{12}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \nonumber \\
    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \nonumber \\
    &\qquad + p_{23}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \nonumber \\
    &\qquad + p_{123}        \cdot\log(1+x'^2-y'^2-x'^2y'^2) \nonumber \\

@@

    &\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \nonumber \\
    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \nonumber \\
    &\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \nonumber \\
    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \nonumber \\
    &\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \nonumber \\
    &\qquad + p_{13}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \nonumber \\
    &\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \nonumber \\
    &\qquad + p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')).

@@

We use the facts that, for $x',y'\in[0,1]$,

@@

1-x'^2+y'^2-x'^2y'^2 & = (1+x')(1-x')(1+y'^2) \le (1+x')(1-x')(1+y') \\
1+x'^2-y'^2-x'^2y'^2 & = (1+x'^2)(1+y')(1-y') \le (1+x')(1+y')(1-y') \\
1+x'^2+y'^2+x'^2y'^2 & = (1+x'^2)(1+y'^2) \le (1+x')(1+y') \\
4x'y' & = 2x' \cdot 2y' \le (1+x'^2)(1+y'^2) \le (1+x')(1+y').

@@

    p_{\emptyset} & = \log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
                  & \le \log(1+x'^2+y'^2+4x'y'+x'^2y'^2) \\
                  & \le \log(2(1+x')(1+y'))

@@

    p_{13} & = \log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
                  & \le \log(1+x'^2+y'^2+x'^2y'^2) \\
                  & \le \log((1+x')(1+y')).

@@

&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\
&\qquad \le      p_{\emptyset}  \cdot\log(2(1+x')(1+y'))
+ p_{1}          \cdot\log((1+x')(1-x')(1+y')) \\
    &\qquad\qquad + p_{2}          \cdot\log((1+x')(1+y')(1-y'))
+ p_{3}          \cdot\log((1+x')(1-x')(1+y')) \\
    &\qquad\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+y')(1-y'))
+ p_{13}         \cdot\log((1+x')(1+y')) \\
    &\qquad\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+y')(1-y'))
+ p_{123}        \cdot\log((1+x')(1+y')(1-y'))

@@

    &\qquad\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2)
+ p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
    &\qquad\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y'))
+ p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
    &\qquad\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y'))
+ p_{13}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
    &\qquad\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y'))
+ p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\

@@

+ \log(1+x')
+ (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
&\qquad\qquad + \log(1+y')
+ (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
&\qquad\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x')
+ (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
&\qquad\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y')
+ (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
&\qquad\qquad + (1-p_{13})\cdot\log(1+w')
+ p_{13}\cdot\log(1-w')

@@

+ (3-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x')
+ 2(p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
&\qquad\qquad + (3-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y')
+ 2(p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\
&\qquad\qquad + (1-p_{13})\cdot\log(1+w')
+ p_{13}\cdot\log(1-w')

@@

&    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\})) \\

@@

+ a_{1}\cdot\log(1+x')
+ a_{2}\cdot\log(1-x')
+ a_{3}\cdot\log(1+y') \\
&\qquad\qquad + a_{4}\cdot\log(1-y')
+ (1-p_{13})\cdot\log(1+w')
+ p_{13}\cdot\log(1-w')

@@

\hat{x}' = \frac{a_{1}-a_{2}}{a_{1}+a_{2}}, \ \hat{y}' = \frac{a_{3}-a_{4}}{a_{3}+a_{4}}.

@@

We check that $0 \le \hat{x}' \le 1$ and $0 \le \hat{y}' \le 1$ so these maximizers are valid.

@@

\hat{x}' = \frac{3-\tilde{p} - 2\tilde{p}}{3-\tilde{p}+2\tilde{p}} = \frac{3-3\tilde{p}}{3+\tilde{p}},

@@

and, since $0 \le \tilde{p} \le 1/2$, $0 \le \hat{x}' \le 1$.
The same approach works for $\hat{y}'$.

@@

&\qquad := L^{1}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).

@@

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
Call
\begin{equation}
    \begin{aligned}
        a_{1}' &= a_{1}-2p_{13}, \\
        a_{2}' &= a_{2}+2p_{13}, \\
        a_{3}' &= a_{3}-2p_{13}, \\
        a_{4}' &= a_{4}+2p_{13}. \\
    \end{aligned}
    \label{eq:a_const_prime}
\end{equation}
We also now need to check that
$$
\hat{x}' = \frac{a_{1}'-a_{2}'}{a_{1}'+a_{2}'}, \ \hat{y}' = \frac{a_{3}'-a_{4}'}{a_{3}'+a_{4}'}
$$
are both between zero and one.
It will suffice to show
$$
0 \le \tilde{p} - 2p_{13} \le 1
$$
and
$$
0 \le \tilde{p} + 2p_{13} \le 1
$$
for $\tilde{p}$ defined above, which is equivalent to showing $p_{13} \le 1/4$.
Using Table~\ref{tab:sitepatprob},
$$
p_{13} = \frac{1}{8}\left(1 + x^2 + y^2 - 4xy^2 + x^2y^2\right).
$$
We see that
$$
1 + x^2 + y^2 - 4xy^2 + x^2y^2 \le 2 \iff y^2\left(1 - 4x + x^2\right) \le 1 - x^2
$$
and
$$
y^2\left(1 - 4x + x^2\right) \le 1 - 4x + x^2 \le 1 - x^2,
$$
the last inequality holding if $x \le 2$.
Thus, $p_{13} \le 2/8 = 1/4$ and our $\hat{x}'$ and $\hat{y}'$ are valid.

>>>>

\begin{table}
\centering
\begin{tabular}{|l|l|l|}
    \hline
$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
    \hline
    $\emptyset$    &$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
    $\{1\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{2\}$        &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    $\{3\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{1,2\}$      &$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,3\}$      &$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
    $\{2,3\}$      &$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,2,3\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    \hline
\end{tabular}
\caption{Site pattern probabilities.
All values are multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
    \multicolumn{3}{c}{$\tau=\tau_1$}\\
    \hline
    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
    \hline
    $\emptyset$&
    $\emptyset$&
    $(1+x)^2   (1+w)(1+y)^2$\\
     $\{1\}$    &
    $\emptyset$&
    $(1+x)(1-x)(1+w)(1+y)^2$\\
     $\{2\}$    &
    $\emptyset$&
    $(1+x)^2   (1+w)(1+y)(1-y)$\\
     $\{3\}$    &
    $\emptyset$&
    $(1+x)(1-x)(1+w)(1+y)^2$\\
    $\{1,2\}$  &
    $\{\emptyset,\{1,2\}\}$&
    $(1+x)(1-x)(1+w)(1+y)(1-y)$\\
    $\{1,3\}$  &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{1\}\\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1-x)^2   (1+w)(1+y)^2\\
                    (1+x)^2   (1-w)(1+y)^2\\
                    (1+x)^2   (1+w)(1-y)^2
                \end{array}$\\
    $\{2,3\}$  &
                $\{\emptyset,\{1,2\}\}$&
                $(1+x)(1-x)(1+w)(1+y)(1-y)$\\
    $\{1,2,3\}$&
                $\{1,2\}$&
                $(1+x)^2   (1+w)(1+y)(1-y)$\\
    \hline
    \multicolumn{3}{c}{$\tau=\tau_2$}\\
    \hline
    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
    \hline
    $\emptyset$       &$\emptyset$&$(1+x)^2   (1+w)(1+y)^2$\\
    $\{1\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                        (1+x)(1-x)(1+w)(1+y)^2\\
                        (1+x)^2   (1-w)(1+y)(1-y)
                    \end{array}$\\
      $\{2\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)^2   (1+w)(1+y)(1-y)\\
                    (1+x)(1-x)(1-w)(1+y)^2
                    \end{array}$\\
      $\{3\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)(1-x)(1+w)(1+y)^2\\
                    (1+x)^2   (1-w)(1+y)(1-y)
                    \end{array}$\\
     $\{1,2\}$         &
    $\left\{\begin{array}{l}
                    \{\emptyset,\{1,2\}\}\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)(1-x)(1+w)(1+y)(1-y)\\
                    (1+x)^2   (1-w)(1+y)^2
                    \end{array}$\\
     $\{1,3\}$         &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{\{1\},\{2\}\}\\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1-x)^2   (1+w)(1+y)^2\\
                    (1+x)(1-x)(1-w)(1+y)(1-y)\\
                    (1+x)^2   (1+w)(1-y)^2
                    \end{array}$\\
      $\{2,3\}$        &
    $\left\{\begin{array}{l}
                    \{\emptyset,\{1,2\}\}\\
                    \{1\}\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)(1-x)(1+w)(1+y)(1-y)\\
                    (1-x)^2   (1-w)(1+y)^2\\
                    (1+x)^2   (1-w)(1-y)^2
                    \end{array}$\\
     $\{1,2,3\}$       &
    $\left\{\begin{array}{l}
                    \{2\} \\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)(1-x)(1-w)(1+y)^2 \\
                    (1+x)^2   (1+w)(1+y)(1-y)
                    \end{array}$\\
    \hline
\end{tabular}
\caption{Likelihood calculations for all site patterns and ancestral state partitions.
All values multiplied by $1/32$.
Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
See Appendix for full calculations.}
\label{tab:likelihoods}
\end{table}

@@

Assume $\tau^*=\tau_1$ and $t^*=\{x^*,y^*,y^*\}$.
Following \eqref{eq:hadamard_probability}, let

@@

p_{\emptyset} := p_{\emptyset}(x^*, y^*) = P(\siteSplitRV=\emptyset \mid \tau=\tau_1,t=t^*),

@@

    \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t))
    &=        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw+x^2y^2) \nonumber \\
    &\qquad + p_{1}          \cdot\log(1-x^2+y^2-x^2y^2) \nonumber \\
    &\qquad + p_{2}          \cdot\log(1+x^2-y^2-x^2y^2) \nonumber \\
    &\qquad + p_{3}          \cdot\log(1-x^2+y^2-x^2y^2) \nonumber \\
    &\qquad + p_{12}         \cdot\log(1-x^2-y^2+x^2y^2) \nonumber \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw+x^2y^2) \nonumber \\
    &\qquad + p_{23}         \cdot\log(1-x^2-y^2+x^2y^2) \nonumber \\
    &\qquad + p_{123}        \cdot\log(1+x^2-y^2-x^2y^2) \nonumber \\

@@

    &\qquad + p_{\emptyset}  \cdot\log((1+x)^2   (1+w)(1+y)^2) \nonumber \\
    &\qquad + p_{1}          \cdot\log((1+x)(1-x)(1+w)(1+y)^2) \nonumber \\
    &\qquad + p_{2}          \cdot\log((1+x)^2   (1+w)(1+y)(1-y)) \nonumber \\
    &\qquad + p_{3}          \cdot\log((1+x)(1-x)(1+w)(1+y)^2) \nonumber \\
    &\qquad + p_{12}         \cdot\log((1+x)(1-x)(1+w)(1+y)(1-y)) \nonumber \\
    &\qquad + p_{13}         \cdot\log((1+x)^2   (1-w)(1+y)^2) \nonumber \\
    &\qquad + p_{23}         \cdot\log((1+x)(1-x)(1+w)(1+y)(1-y)) \nonumber \\
    &\qquad + p_{123}        \cdot\log((1+x)^2   (1+w)(1+y)(1-y)).

@@

We use the facts that, for $x,y\in[0,1]$,

@@

1-x^2+y^2-x^2y^2 & = (1+x)(1-x)(1+y^2) \le (1+x)(1-x)(1+y) \\
1+x^2-y^2-x^2y^2 & = (1+x^2)(1+y)(1-y) \le (1+x)(1+y)(1-y) \\
1+x^2+y^2+x^2y^2 & = (1+x^2)(1+y^2) \le (1+x)(1+y) \\
4xy & = 2x \cdot 2y \le (1+x^2)(1+y^2) \le (1+x)(1+y).

@@

    p_{\emptyset} & = \log(1+x^2+y^2+4xyw+x^2y^2) \\
                  & \le \log(1+x^2+y^2+4xy+x^2y^2) \\
                  & \le \log(2(1+x)(1+y))

@@

    p_{13} & = \log(1+x^2+y^2-4xyw+x^2y^2) \\
                  & \le \log(1+x^2+y^2+x^2y^2) \\
                  & \le \log((1+x)(1+y)).

@@

&    \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \\
&\qquad \le      p_{\emptyset}  \cdot\log(2(1+x)(1+y))
+ p_{1}          \cdot\log((1+x)(1-x)(1+y)) \\
    &\qquad\qquad + p_{2}          \cdot\log((1+x)(1+y)(1-y))
+ p_{3}          \cdot\log((1+x)(1-x)(1+y)) \\
    &\qquad\qquad + p_{12}         \cdot\log((1+x)(1-x)(1+y)(1-y))
+ p_{13}         \cdot\log((1+x)(1+y)) \\
    &\qquad\qquad + p_{23}         \cdot\log((1+x)(1-x)(1+y)(1-y))
+ p_{123}        \cdot\log((1+x)(1+y)(1-y))

@@

    &\qquad\qquad + p_{\emptyset}  \cdot\log((1+x)^2   (1+w)(1+y)^2)
+ p_{1}          \cdot\log((1+x)(1-x)(1+w)(1+y)^2) \\
    &\qquad\qquad + p_{2}          \cdot\log((1+x)^2   (1+w)(1+y)(1-y))
+ p_{3}          \cdot\log((1+x)(1-x)(1+w)(1+y)^2) \\
    &\qquad\qquad + p_{12}         \cdot\log((1+x)(1-x)(1+w)(1+y)(1-y))
+ p_{13}         \cdot\log((1+x)^2   (1-w)(1+y)^2) \\
    &\qquad\qquad + p_{23}         \cdot\log((1+x)(1-x)(1+w)(1+y)(1-y))
+ p_{123}        \cdot\log((1+x)^2   (1+w)(1+y)(1-y)) \\

@@

+ \log(1+x)
+ (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x) \\
&\qquad\qquad + \log(1+y)
+ (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y) \\
&\qquad\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x)
+ (p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x) \\
&\qquad\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y)
+ (p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y) \\
&\qquad\qquad + (1-p_{13})\cdot\log(1+w)
+ p_{13}\cdot\log(1-w)

@@

+ (3-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x)
+ 2(p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x) \\
&\qquad\qquad + (3-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y)
+ 2(p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y) \\
&\qquad\qquad + (1-p_{13})\cdot\log(1+w)
+ p_{13}\cdot\log(1-w)

@@

&    \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \\

@@

+ a_{1}\cdot\log(1+x)
+ a_{2}\cdot\log(1-x)
+ a_{3}\cdot\log(1+y) \\
&\qquad\qquad + a_{4}\cdot\log(1-y)
+ (1-p_{13})\cdot\log(1+w)
+ p_{13}\cdot\log(1-w)

@@

\hat{x} = \frac{a_{1}-a_{2}}{a_{1}+a_{2}}, \ \hat{y} = \frac{a_{3}-a_{4}}{a_{3}+a_{4}}, \ \hat{w} = 1-2p_{13}.

@@

Trivially $0 \le \hat{w} \le 1$.
We check that $0 \le \hat{x} \le 1$ and $0 \le \hat{y} \le 1$ so these maximizers are valid.

@@

\hat{x} = \frac{3-\tilde{p} - 2\tilde{p}}{3-\tilde{p}+2\tilde{p}} = \frac{3-3\tilde{p}}{3+\tilde{p}},

@@

and, since $0 \le \tilde{p} \le 1/2$, $0 \le \hat{x} \le 1$.
The same approach works for $\hat{y}$.

@@

&\qquad := L^{1}_{\tau_1,\tau_1}(x^*, y^*).

@@

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
Call
\begin{equation}
    \begin{aligned}
        a_{1}' &= a_{1}-2p_{13}, \\
        a_{2}' &= a_{2}+2p_{13}, \\
        a_{3}' &= a_{3}-2p_{13}, \\
        a_{4}' &= a_{4}+2p_{13}. \\
    \end{aligned}
    \label{eq:a_const_prime}
\end{equation}
We also now need to check that
$$
\hat{x} = \frac{a_{1}'-a_{2}'}{a_{1}'+a_{2}'}, \ \hat{y} = \frac{a_{3}'-a_{4}'}{a_{3}'+a_{4}'}
$$
are both between zero and one.
It will suffice to show
$$
0 \le \tilde{p} - 2p_{13} \le 1
$$
and
$$
0 \le \tilde{p} + 2p_{13} \le 1
$$
for $\tilde{p}$ defined above, which is equivalent to showing $p_{13} \le 1/4$.
Using Table~\ref{tab:sitepatprob} and $t^*$,
$$
p_{13} = \frac{1}{8}\left(1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2\right).
$$
We see that
$$
1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 2 \iff (y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - (x^*)^2
$$
and
$$
(y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - 4x^* + (x^*)^2 \le 1 - (x^*)^2,
$$
the last inequality holding if $x^* \le 2$.
Thus, $p_{13} \le 2/8 = 1/4$ and our $\hat{x}$ and $\hat{y}$ are valid.

To bound the case of $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$,
\begin{align*}
&    \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'}
+ a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
&\qquad\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}}
+ a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
+ \log(2)
- \log(8) \\
&\qquad := L^{2}_{\tau_1,\tau_1}(x^*, y^*),
\end{align*}
and, for the case of $\ancestralSplitPartition_j(\tau_1, t) = \{1,2\}$,
\begin{align*}
&    \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}}
+ a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
&\qquad\qquad+ a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'}
+ a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'}
+ \log(2)
- \log(8) \\
&\qquad := L^{3}_{\tau_1,\tau_1}(x^*, y^*).
\end{align*}
All bounds are functions only of $x,y$ through the true generating probabilities $x^*, y^*$.
Call the upper bound
$$
L_0(x^*, y^*) := \max(L^{1}_{\tau_1,\tau_1}(x^*, y^*), L^{2}_{\tau_1,\tau_1}(x^*, y^*), L^{3}_{\tau_1,\tau_1}(x^*, y^*)).
$$
